Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-11 _tex/paper2.tex
Built site for gh-pages

However, because CJ experiments rarely use the exhaustive pairing of
sampled judges, stimuli, and individuals--due to concerns about the
practical feasibility of the comparison task
\citep{Boonen_et_al_2020}--a realistic scenario must also account for
judges comparing only specific stimuli from certain individuals.

@@

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
directly access this outcome due to the comparison mechanism. Moreover,
it depicts \(C\) as a causal factor influencing \(O_{R}\). As with the
sampling mechanism, a square encloses \(C\) to indicate that it is a
conditioned variable--that is, one that determines \emph{which repeated
judgments the subset of judges makes for the subset of stimuli produced
by the sampled individuals}. In essence, \(C\) reflects the assumption
that judges \emph{do not} perform all possible repeated judgments for
the selected stimuli. Instead, they carry out a ``sufficient number of
comparisons'' \(n_{C}\) to allow for accurate estimation of the
proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline below. First, we assume that \(O_{R}\) follows a
Bernoulli distribution\footnote{The binomial distribution--including its
  special case, the Bernoulli distribution-- represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that the probability
distributions for the sample and comparison mechanisms, \(P(S)\) and
\(P(C)\), included in the probabilistic model \ref{fig-cj16_prob}, are
omitted from the statistical model \ref{fig-cj16_stat}. We exclude these
probabilities from the statistical definition because, although they are
part of the joint distribution of the data, they do not influence the
dependence between \(O_{R}\) and \(D_{R}\), since \(S\) and \(C\) are
governed by random selection designs\footnote{Randomization ensures that
  data---and, by extension, an estimator---satisfies several key
  identification properties, such as common support, no interference,
  and consistency. The most critical property, however, is the
  elimination of confounding. \emph{Confounding} occurs when an external
  variable, such as \(X_{I}\), simultaneously influences both the
  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
processes represent the underlying written-quality traits of the texts.
Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). For clarity, the
square brackets in \(D_{R}\) only indicate the relevant indices within
each trait vector; they do not represent any data subsetting.

@@

Third, we define \(T_{IA}\) as a linear combination of the students'
underlying writing-quality traits \(T_{I}\), the effects of relevant
text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
each element of \(e_{IA}\) is independent and identically distributed
\((iid)\) according to a Normal distribution, with a mean of zero and a
standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
proportion of 1 to establish the scale of the latent trait, which is
required in any latent variable model \citep{Depaoli_2021}. Note that
because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
normally distributed. This result indicates that our model reverts to
the original assumptions for the discriminal processes of stimuli
proposed by \citet{Thurstone_1927b} (see
Table~\ref{tbl-thurstone_cases}).

@@

Sixth, we define \(B_{J}\) as a linear combination of the effects of
relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
which each judge belongs. For instance, if \(Z_{J} = 1, \dots, 3\)
represents three groups of judges with varying expertise, the model
imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
scale of the latent trait \citep{Depaoli_2021}.

>>>>

However, due to concerns about the practical feasibility of the
comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
implement an exhaustive pairings of sampled judges, stimuli, and
individuals. Thus, a realistic scenario must account for the fact that
judges typically compare only a subset of stimuli authored by certain
individuals.

@@

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
directly access to \(O^{sc}_{R}\) due to the comparison mechanism.
Moreover, the DAG depicts \(C\) as a causal factor influencing
\(O_{R}\), enclosing it in a square to indicate that it is a conditioned
variable. Specifically, \(C\) determines \emph{which repeated judgments
the subset of judges makes for the subset of stimuli produced by the
sampled individuals}. In essence, \(C\) reflects the assumption that
judges \emph{do not} perform all possible repeated judgments for the
selected stimuli. Instead, they carry out a ``sufficient number of
comparisons'' \(n_{C}\) to allow for accurate estimation of the
proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
  binomial distribution--including its special case, the Bernoulli
  distribution-- represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that when only two
  un-ordered outcomes are possible, and the expected numbers of each
  type of event are assumed to be constant, then the binomial
  distribution is the one that is most consistent with these constraints
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that while the
probabilistic model \ref{fig-cj16_prob} includes the probability
distributions for the sampling and comparison mechanisms, \(P(S)\) and
\(P(C)\), we omit these from the statistical model \ref{fig-cj16_stat}.
We do so because, although these distributions contribute to the joint
distribution of the data, they do not affect the relationship between
\(O_{R}\) and \(D_{R}\). This lack of influence is due to the random
selection designs that govern \(S\) and \(C\)\footnote{Randomization
  ensures that data---and, by extension, an estimator---satisfies
  several key identification properties, such as common support, no
  interference, and consistency. The most critical property, however, is
  the elimination of confounding. \emph{Confounding} occurs when an
  external variable, such as \(X_{I}\), simultaneously influences both
  the outcome (e.g., \(O_{R}\)) and a variable of interest (e.g.,
  \(S\)), resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
processes represent the underlying written-quality trait of the texts.
Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). It is important
to clarify that the square brackets in \(D_{R}\) merely indicate the
relevant indices for each trait vector, without implying any subsetting
of the data.

@@

Third, we define \(T_{IA}\) as a linear combination of the students'
underlying writing-quality traits \(T_{I}\), the effects of relevant
text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
each element of \(e_{IA}\) is independent and identically distributed
\((iid)\) according to a Normal distribution, with a mean of zero and a
standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
proportion of 1 to establish the scale of the latent trait, which is
required in any latent variable model \citep{Depaoli_2021}. Note that
because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
normally distributed. This result implies that our model reverts to
assuming Normal discriminal processes for the stimuli, as originally
proposed by \citet{Thurstone_1927b} (see
Table~\ref{tbl-thurstone_cases}).

@@

Sixth, we define \(B_{J}\) as a linear combination of the effects of
relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
which each judge belongs. For instance, if \(Z_{J}\) is a variable that
can take on three values representing three groups of judges with
varying expertise, the model imposes the constraint
\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
\citep{Depaoli_2021}.
